Best Practices for Developing Sustainable Scientific Software

### 2016-05-16 WSSSPE3/best_practices/main.tex
Update main.tex

>>>>

Scientific software is an integral part of research in a plethora of areas such as applied natural sciences, science of management and humanities - not only increasing the efficiency to find answers to research questions but also enhancing methods to answer some research questions at all. While the importance of developing scientific software is evident in the large number of software package, methods and algorithms applied in science, sustaining it and thus make it easily usable and re-usable for the user communities entails challenges immanent in being part of the highly dynamic environment in research. 
Reviewing multiple past articles and talks at different meetings like WSSSPEx
\cite{WSSSPE1, WSSSPE2, 5069157, Blatt_WSSSPE, Ahern_WSSSPE} and analyzing and
promoting sustainable scientific software makes it clear that there are several
common and recurring ideas that underpin success in developing sustainable
software. However, outside of a small community, this knowledge is not widely
shared. This is especially true for the large community of scientists who
generate most of the software used by scientists but are not primarily software
developers. In this scenario, a clear and precise exposition of these best
practices collected from many sources and open collaboration among all in the
community in a single source (e.g., journal paper, tutorial) that can be widely
disseminated is necessary and likely to be very valuable.

Core questions that will need to be explored are in reproducibility, 
reliability,  usability, extensibility, knowledge management, as well as continuity
(transitions between people). Answers to these questions will guide us on how a software
tool becomes part of the core workflow of well identified users (stakeholders)
relating to tool success and hence sustainability.
%\katznote{prev sentence is complex and awkward} % see fix
Ideas  that may need to be explored include:
\begin{itemize}
\item Requirements engineering to create tools with immediate uptake;
\item When should software ``die''?
\item Catering to disruptive developments in environment (e.g., new hardware,
new methodology);
\item Dimensions of sustainability: economic, technical, environmental and
obsolescence. 
%\katznote{not sure what the prev. comment goes with} %social.
\end{itemize}

Sustainability requires community participation in code development and/or a
wide adoption of software. The larger the community base is using a piece of
software, the better are the funding possibilities and thus also the
sustainability options. Additionally developer commitment to an application is
essential and experience shows that software packages with an evangelist
imposing strong inspiration and discipline are more likely to achieve
sustainability. While a single person can push sustainability to a certain
level, open source software also needs sustained commitment from the developer
community. Such sustained commitments include diverse tasks and roles, which can
be fulfilled by diverse developers with different knowledge levels. Besides
developing software and appropriate software management with measures for
extensibility and scalability of the software, active (expertise) support for
users via a user forum with a quick turnaround is crucial. The barrier to entry
for the community as users as well as developers has to be as low as possible.
